Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
add sstimer

### Other storage approaches.

?? do we want to report any other approaches here?
In particular, should we look at minpack.lm to see how the Marquardt stabilization 
is applied? This is a very complicated program, since it calls C to call Fortran,
and the Fortran is layered in a way that each part is well-documented but the
general flow and linkage of the parts is non-obvious.

## Sequential or full Jacobian computation

We could compute a row of the Jacobian plus the corresponding residual element
and process this before computing the next row etc. This means the full Jacobian
does not have to be stored. In Nashlib, Algorithms 3 and 4, we used row-wise data 
entry in linear least squares via Givens' triangularization (QR decompostition),
with the possibility of extending the QR to a singular value decomposition.
Forming the SSCP matrix can also be generated row-wise as well. We do not 
anticipate that any of these approaches will offer advantages that are so
great that we would choose them.

@@

The `nlsr` package, however, attempts to compute analytic derivatives using symbolic
and analytic derivative possibilities from other parts of R. A great advantage of
these is that they are for a given point, and do not suffer the issue that a step
must be taken, thereby chancing the violation of a constraint or else an inadmissible
set of arguments for special functions. In my experience, the use of analytic derivatives
is not much faster than the finite difference approximations, and generally finite
difference methods iterate as quickly as the "exact" derivative methods to the 
neighbourhood of a solution. It is in allowing good estimates of convergence 
criteria that the analytic derivatives seem to show their advantage i.e., knowing
we are at a solution rather than the speed in getting there.

@@

R allows the nonlinear least squares problem to be presented via a formula
for the model. This is generally very attractive to users, since they can
specify their problem in ways that often accord with how they think about them.
Many (most?) researchers do not write program code, so preparing residual and
Jacobian functions is foreign to them.

@@

- translating a formula to a function for its actual evaluation may be tricky,
  since we must essentially parse the formula. R has some tools for this that
  we use.

@@

- checking for correctness of inputs and translations can be considered to
  take up most of the code in a reliable and robust package.

@@

- use a row-wise generation of the Jacobian in either a Givens' QR or SSCP approach.
  This saves space for the Jacobian as well as as well as the working matrices of
  the Gauss-Newton or Marquardt iterations;

- if the number of parameters to estimate is large enough, then a normal equations
  approach using a compact storage of the lower triangle of the SSCP matrix. 
  However, the scale of the saving is really very small in comparison to 
  the size of most programs.

@@

Do we plan to use only timing? Or size of code and memory used? ??

@@

>>>>

<!-- ### Other storage approaches. -->

<!-- ?? do we want to report any other approaches here? -->
<!-- In particular, should we look at minpack.lm to see how the Marquardt stabilization  -->
<!-- is applied? This is a very complicated program, since it calls C to call Fortran, -->
<!-- and the Fortran is layered in a way that each part is well-documented but the -->
<!-- general flow and linkage of the parts is non-obvious. -->

### Sequential or full Jacobian computation

It is possible to compute a row of the Jacobian plus the corresponding residual element
and process these, for example, into the SSCP matrix $J^t J$ and gradient $- J^t r$  before computing 
the next row etc. This means the full Jacobian does not have to be stored. 
In Nashlib, Algorithms 3 and 4, we used row-wise data 
entry in **linear** least squares via Givens' triangularization (QR decompostition),
with the possibility of extending the QR to a singular value decomposition.
While such approaches were forced upon us in the early 1970s when they were devised,
we do not anticipate that any of these approaches will now offer enough advantage that
we would use them as a general choice. They may, however, be appropriate for some 
problems where the number of observations is extremely large.

@@

The `nlsr` package, however, attempts to compute analytic derivatives using symbolic
and analytic derivative possibilities from other parts of R. A great advantage of
these is that they are for a given point, and do not suffer the issue that a step
must be taken, thereby chancing the violation of a constraint or else an inadmissible
set of arguments for special functions. In my experience, the use of analytic derivatives
does **not** give a much faster solution than the finite difference approximations.
Indeed, finite difference methods iterate as quickly as the "exact" derivative methods to the 
neighbourhood of a solution. This is much along the same lines as secant versus Newton
methods for solving nonlinear equations. Analytic derivatives do, however, show their
advantage in allowing us to more reliably compute convergence 
criteria. That is, we know we are at a solution and do not attempt further iterations.

@@

R allows the nonlinear least squares problem to be presented via a **formula**
for the model. This is generally very attractive to users, since they can
specify their problem in ways that often accord with how they think about them.
Many, possibly most, researchers do not write program code, so preparing residual and
Jacobian functions is foreign to them.

@@

- translating a formula to a function for its actual evaluation may be tricky,
  since we must essentially parse the formula. R has some tools for this that
  we use. Nevertheless, the automated conversion may risk setting up calculations
  in ways that could incur ill-conditioning.

@@

- checking for correctness of inputs and translations may
  take up most of the code in a reliable and robust package.

@@

- use a row-wise generation of the Jacobian in either a Givens' QR or SSCP approach, 
  as mentioned above. 
  This saves space for the Jacobian as well as as well as some of the working storage of
  the Gauss-Newton or Marquardt iterations;

- if the number of parameters to estimate is large enough, then a normal equations
  approach using a compact storage of the lower triangle of the SSCP matrix may help. 
  However, the scale of the saving is really very small in comparison to 
  the size of most programs.

@@

Most workers report execution timings as the principal measure of performance.
This is generally acceptable for comparing programs, and we often then attribute
the results to the underlying methods, though clearly programming choices can have
a quite large effect.

For example, we can form sums of squares in R using `for` loops, a vectorized
`sum()` or the `crossprod()` function. These have different performance profiles.
Let us form the sums of squares of different length random vectors.

```
 n        t(forloop) : ratio          t(sum) : ratio       t(crossprod)    all.equal 
100           31701  :  2.1996        15541  :  1.0784           14412       TRUE 
1000          35604  :  13.004       2900.1  :  1.0592            2737.9     TRUE 
10000        323969  :  27.826        21051  :  1.8081           11643       TRUE 
1e+05       3120277  :  30.604       291234  :  2.8564          101957       TRUE 
1e+06      31140093  :  29.231      4701221  :  4.4131         1065296       TRUE 
1e+07     317191471  :  21.776     40540123  :  2.7832        14565799       TRUE 
```

These are averages of 100 executions of each sum, and we would expect them to be
stable, but there is some variation, likely due to other processes running in an
active computer, as seen by rerunning the script `sstimer.R`.

```
 n        t(forloop) : ratio          t(sum) : ratio        t(crossprod)   all.equal 
100           35643  :  2.4688        13258  :  0.91831           14437      TRUE 
1000          33172  :  12.142         3689  :  1.3503             2731.9    TRUE 
10000        324269  :  27.958        28028  :  2.4166            11598      TRUE 
1e+05       3132152  :  30.765       304086  :  2.9868           101810      TRUE 
1e+06      31262348  :  28.846      3759337  :  3.4688          1083772      TRUE 
1e+07     313522217  :  21.538     38746311  :  2.6618         14556401      TRUE 
```

We also note the very strange result for `n=1000`. ?? No explanation.
Moreover, results in a Windows 10 virtual machine (under VirtualBox)
give the results

```
 n      t(forloop) : ratio            t(sum) : ratio        t(crossprod)   all.equal 
100         43781  :  2.592126        14442  :  0.8550622         16890      TRUE 
1000        34241  :  12.59323         2452  :  0.9018021          2719      TRUE 
10000      348003  :  19.6058         15038  :  0.8472113         17750      TRUE 
1e+05     3404639  :  20.32475       293930  :  1.75468           67512      TRUE 
1e+06    34379874  :  19.36951      3571310  :  2.012065        1774948      TRUE 
1e+07   376569416  :  17.45253     36871902  :  1.70887        21576778      TRUE 
```

It is also possible to measure the computational effort in terms of Jacobian and residual
evaluations, as well as the number of "iterations" used to achieve a solution. The
Hobbs timing above comparing `nlxb()` and `nlxbx()` (i.e., scaled and unscaled Marquardt
approaches) show different numbers of Jacobian computations, which for these methods are
the same as the number of major iterations.

In situations where storage is critical, or for gauging the complexity of code, it may
be useful to report program size or number of lines of code. These measures are all
influenced by comments, program structure, and computing environment issues.

@@

It is worth including some sample test problems in this article to illustrate
the discussion.

?? here or later
